---
title: "wind: wORKFLOW FOR PiRNAs AnD BEYONd"
subtitle: "Comparison of resulted Differential Expressed miRNAs with public results of  GSE68246, regarding Human Breast MCF-7 Cell Line with Cancer Stem Cell Properties"
author: "Constantinos Yeles (Konstantinos Geles)"
date: "`r format(Sys.time(), '%a %b %d %Y')`"
output:
  pdf_document:
    toc: yes
    toc_depth: 3
  html_document:
    toc: yes
    toc_depth: 3
    theme: paper 
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Introduction
Following the differential expression step I will compare the published results of the study with the results of WIND.
We have to take into consideration that different miRNA databases version have been used.

## Load libraries
```{r load libraries, }
suppressPackageStartupMessages({
  library('tidyverse') 
})
```

## Import the public results
```{r public results, }
public_miRNA_down <- readxl::read_excel(path = "Datasets_analysis/GSE68246/DEA_Breast_Cancer_CSC_GRCh38_08_Mar_2021/Public_results_GSE68246.xlsx", sheet = "Table S3 down", skip = 1) %>% 
  select(pub_miRNA = "MCF7 Parental - Small RNA sample-1  mcf7 parental size grouped on mature - Name",
         Sequence = "Feature ID", 
         Fold_Change = "Experiment - Fold Change (original values)",
         FDR = "Kal's Z-test: MCF7 Primary S vs MCF7 Parental original values - FDR p-value correction",
        Bonferroni = "Kal's Z-test: MCF7 Primary S vs MCF7 Parental original values - Bonferroni")

public_miRNA_up <- readxl::read_excel(path = "Datasets_analysis/GSE68246/DEA_Breast_Cancer_CSC_GRCh38_08_Mar_2021/Public_results_GSE68246.xlsx", sheet = "Table S3 up", skip = 1) %>% 
  select(pub_miRNA = "MCF7 Parental - Small RNA sample-1  mcf7 parental size grouped on mature - Name",
         Sequence = "Feature ID", 
         Fold_Change = "Experiment - Fold Change (original values)",
         FDR = "Kal's Z-test: MCF7 Primary S vs MCF7 Parental original values - FDR p-value correction",
        Bonferroni = "Kal's Z-test: MCF7 Primary S vs MCF7 Parental original values - Bonferroni")

public_miRNA <- bind_rows(public_miRNA_down, public_miRNA_up)

rm(public_miRNA_down, public_miRNA_up)
```

## Import the wind results
```{r identified DE, }
wind_res <- read_tsv("Datasets_analysis/GSE68246/DEA_Breast_Cancer_CSC_GRCh38_08_Mar_2021/all_comparisons_voom_QW_TMM_salmon_fc_LFCs_08_Mar_2021.txt")

# make long format 
wind_res_long <- wind_res %>% 
  pivot_longer(
    cols = -c(smallRNA, gene_type),
    names_to = c( "stat", "contrast", "tool"),
    names_pattern = "^(logFC|adj.P.Val|P.Value)_(.+)_(salmon|FC)"
               ) %>% 
  select(-contrast ) %>% 
  filter(stat != "P.Value") %>% 
  pivot_wider(names_from = c(stat)) 

# find how many DEs per tool
wind_res_long %>% 
  filter(adj.P.Val < 0.05) %>% count(tool)

wind_res_long %>% 
  filter(adj.P.Val < 0.05) %>% 
  count(tool, gene_type, sort = TRUE) %>% 
  pivot_wider(names_from = tool, values_from = n) 
```

## import the sequences from the GTF file
```{r}
smallRNAs_gtf <- file.path("human_data", "sncRNA_piRNBnk_RNACent_GRCh38_v34.gtf") %>% 
  plyranges::read_gff2() %>% 
  as_tibble() %>% 
  select(gene_id,  seq_RNA, gene_type) %>% 
  distinct(gene_id, .keep_all = TRUE) %>% 
  dplyr::rename("smallRNA" = gene_id)
```

## join them with wind results
```{r}
wind_res_long <- wind_res_long %>% 
  left_join(smallRNAs_gtf)
```

## combine with public results
```{r}
wind_plus_pub <- wind_res_long %>% 
  full_join(public_miRNA, by = c("seq_RNA" = "Sequence")) %>% 
  filter(!is.na(pub_miRNA))
```

## search the results
```{r}
wind_plus_pub %>% 
  filter(adj.P.Val < 0.05) %>% 
  count(tool, gene_type, sort = TRUE)
```


